Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-06-25 manuscript/manuscript.Rmd
update ms. and references

Even though EHA is a widely used statistical tool and there already exist many excellent reviews [e.g., @singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @blossfeldTechniquesEventHistory2002; @hosmerAppliedSurvivalAnalysis2011; @teachmanAnalyzingSocialProcesses1983; @box-steffensmeierEventHistoryModeling2004] and tutorials [e.g., @allisonSurvivalAnalysisUsing2010; @landesIntroductionEventHistory2020], we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain **how** to do such analyses in the context of experimental psychology, rather than repeat in any detail **why** you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations (Allison, 2010). In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010; @gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019]. We also visualize and discuss the types of time-to-event data that are obtained in typical RT tasks and supply relevant regression equations in sections A and E of the Supplemental Material, respectively.

@@

To avoid possible confusion in terminology used, it is worth noting that EHA is known by various labels, such as survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis (Singer & Willett, 2003). In this paper, we choose to use the term EHA throughout.

@@

There are different types of EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units. 
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper. 
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable. 
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well (REFS). 
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than ~200 trials per condition per experiment [@panisAnalyzingResponseTimes2020].
Moreover, as indicated by @allisonSurvivalAnalysisUsing2010, learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

>>>>

Even though EHA is a widely used statistical tool and there already exist many excellent reviews [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @blossfeldTechniquesEventHistory2002; @hosmerAppliedSurvivalAnalysis2011; @teachmanAnalyzingSocialProcesses1983; @box-steffensmeierEventHistoryModeling2004; @millsIntroducingSurvivalEvent2011] and tutorials [@allisonSurvivalAnalysisUsing2010; @landesIntroductionEventHistory2020; @elmerModelingCategoricalTimetoevent2023; @lougheedMultilevelSurvivalAnalysis2019; @stoolmillerIntroductionUsingMultivariate2015; @stoolmillerModelingHeterogeneitySocial2006], we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain **how** to do such analyses in the context of experimental psychology, rather than repeat in any detail **why** you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

EHA is a class of statistical approaches to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations [@allisonSurvivalAnalysisUsing2010]. In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010; @gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019]. We also visualize and discuss the types of time-to-event data that are obtained in typical RT tasks and supply relevant regression equations in sections A and E of the Supplemental Material, respectively.

@@

To avoid possible confusion in terminology used, it is worth noting that EHA is known by various labels, such as survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis [@singerAppliedLongitudinalData2003]. In this paper, we choose to use the term EHA throughout.

@@

There are different types of EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units. 
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper. 
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable. 
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well [@bloxomEstimatingResponseTime1984; @luceResponseTimesTheir1991; @vanzandtHowFitResponse2000]. 
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than ~200 trials per condition per experiment [@panisAnalyzingResponseTimes2020].
Moreover, as indicated by @allisonSurvivalAnalysisUsing2010, learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

### 2025-06-25 manuscript/manuscript.tex
update ms. and references

Even though EHA is a widely used statistical tool and there already exist many excellent reviews (Allison, 1982; Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; e.g., Singer \& Willett, 2003; Teachman, 1983) and tutorials (e.g., Allison, 2010; Landes, Engelhardt, \& Pelletier, 2020), we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain \textbf{how} to do such analyses in the context of experimental psychology, rather than repeat in any detail \textbf{why} you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

There are different types of EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper.
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable.
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well (REFS).
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than \textasciitilde200 trials per condition per experiment (REF PANIS).
Moreover, as indicated by Allison (2010), learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

@@

@@

@@

@@

@@

@@

>>>>

Even though EHA is a widely used statistical tool and there already exist many excellent reviews (Allison, 1982; Blossfeld \& Rohwer, 2002; Box-Steffensmeier, 2004; Hosmer, Lemeshow, \& May, 2011; Mills, 2011; Singer \& Willett, 2003; Teachman, 1983) and tutorials (Allison, 2010; Elmer, Van Duijn, Ram, \& Bringmann, 2023; Landes, Engelhardt, \& Pelletier, 2020; Lougheed, Benson, Cole, \& Ram, 2019; Stoolmiller, 2015; Stoolmiller \& Snyder, 2006), we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain \textbf{how} to do such analyses in the context of experimental psychology, rather than repeat in any detail \textbf{why} you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

@@

There are different types of EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.
As a lab, and mainly for practical reasons, we have much more experience using discrete-time EHA, and that is the approach that we describe and focus on in this paper.
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable.
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well (Bloxom, 1984; Luce, 1991; Van Zandt, 2000).
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological RT data sets for which there are less than \textasciitilde200 trials per condition per experiment (Panis, Schmidt, et al., 2020).
Moreover, as indicated by Allison (2010), learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

@@

\bibitem[\citeproctext]{ref-bloxomEstimatingResponseTime1984}
Bloxom, B. (1984). Estimating response time hazard functions: {An} exposition and extension. \emph{Journal of Mathematical Psychology}, \emph{28}(4), 401--420. \url{https://doi.org/10.1016/0022-2496(84)90008-7}

@@

\bibitem[\citeproctext]{ref-elmerModelingCategoricalTimetoevent2023}
Elmer, T., Van Duijn, M. A. J., Ram, N., \& Bringmann, L. F. (2023). Modeling categorical time-to-event data: {The} example of social interaction dynamics captured with event-contingent experience sampling methods. \emph{Psychological Methods}. \url{https://doi.org/10.1037/met0000598}

@@

\bibitem[\citeproctext]{ref-lougheedMultilevelSurvivalAnalysis2019}
Lougheed, J. P., Benson, L., Cole, P. M., \& Ram, N. (2019). Multilevel survival analysis: {Studying} the timing of children's recurring behaviors. \emph{Developmental Psychology}, \emph{55}(1), 53--65. \url{https://doi.org/10.1037/dev0000619}

@@

\bibitem[\citeproctext]{ref-millsIntroducingSurvivalEvent2011}
Mills, M. (2011). \emph{Introducing {Survival} and {Event History Analysis}}. 1 Oliver's Yard,~55 City Road,~London~EC1Y 1SP~United Kingdom: SAGE Publications Ltd. \url{https://doi.org/10.4135/9781446268360}

@@

\bibitem[\citeproctext]{ref-stoolmillerIntroductionUsingMultivariate2015}
Stoolmiller, M. (2015). \emph{An {Introduction} to {Using Multivariate Multilevel Survival Analysis} to {Study Coercive Family Process}} (Vol. 1; T. J. Dishion \& J. Snyder, Eds.). Oxford University Press. \url{https://doi.org/10.1093/oxfordhb/9780199324552.013.27}

\bibitem[\citeproctext]{ref-stoolmillerModelingHeterogeneitySocial2006}
Stoolmiller, M., \& Snyder, J. (2006). Modeling heterogeneity in social interaction processes using multilevel survival analysis. \emph{Psychological Methods}, \emph{11}(2), 164--177. \url{https://doi.org/10.1037/1082-989X.11.2.164}

@@

\bibitem[\citeproctext]{ref-vanzandtHowFitResponse2000}
Van Zandt, T. (2000). How to fit a response time distribution. \emph{Psychonomic Bulletin \& Review}, \emph{7}(3), 424--465. \url{https://doi.org/10.3758/BF03214357}
